---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---
## Loading and preprocessing the data
```{r}
## Unzipping the data, loading it into a data frame.
unzip("activity.zip")
data <- read.csv(file = "activity.csv", header = TRUE)

## Tries to remove the file from the hard disk after loading it. Returns a
## boolean value indicating wether it succesfully executed.
file.remove("activity.csv")

## Transforming days (as factors) into time variables - this does not take
## into account the time of day.
data$date <- strptime(data$date, format="%Y-%m-%d")
```

## What is mean total number of steps taken per day?
```{r}
## Define a new sum function that already removes the NA values.
newsum <- function(x){ sum(x,na.rm=TRUE)}

## This function computes the total number of steps on a data frame containing 
## step count information in the format [date, number of steps,interval,...]
totalSteps <- function(inputData){
      ## Groups dates into factors by days.
      stepsgroupedByDay <- split(data$steps,cut(inputData$date,breaks="day"))
      
      ## To compute the total number of steps per day we can split the number
      ## of steps according to the date labels and apply the newsum function.
      totalStepsPerDay <- sapply(stepsgroupedByDay, newsum)
}

totalStepValues <- totalSteps(data)
hist(totalStepValues)
mean(totalStepValues)
median(totalStepValues)
```


## What is the average daily activity pattern?
```{r}
## Note that the interval format is just recording the digits appearing in a
## digital clock, i.e. "1605" corresponds to 4.05 pm or 16:05

## Define a short hand version of the function mean(x, na.rm = TRUE)
newmean <- function(x){mean(x,na.rm =TRUE)}

## This function computes the activity pattern on a data frame containing 
## step count information in the format [date, number of steps,interval,...]
activityPattern <- function(inputData){
      splitData <- split(inputData$steps,inputData$interval)
      avgActivityPerDay <-  sapply(splitData,newmean)
      time <- as.numeric(names(avgActivityPerDay))
      hour <- as.character( (time - time%%100)/100 )
      minute<- as.character(time %%100)
      timeStamps <- paste(hour,minute,"00", sep =":")
      timeOfDay <- strptime(timeStamps,format="%H:%M:%S")
      list(timeOfDay = timeOfDay, averageActivityPerDay = avgActivityPerDay,
           times=time)
}

## averageActivityPerDay contains the average number of steps, over 5 minute
## intervals, across all of the days in the data set.
computedActivity <- activityPattern(data)
averageActivityPerDay <-  computedActivity$averageActivityPerDay
times <- computedActivity$times
timeOfDay <- computedActivity$timeOfDay

### Constructing a plot of the averageActivityPerDay vs. the time of day.

plot(timeOfDay,averageActivityPerDay, type ="l",xlab ="Time of Day",ylab="Average Steps per Day")


### Determining the interval(s) containing the largest average steps per day.

highestAvgSteps <- max(averageActivityPerDay)
times[averageActivityPerDay==highestAvgSteps]
```



## Inputing missing values
```{r}
### Counting the total number of missing values.
## Replacing NAs with the 5-min time-interval average corresponding to 
## the missing value.
boolNaIndices <- is.na(data$steps)
totalNAs <- sum( boolNaIndices )
```

### Replacing the missing values with the average value for the corresponding interval.
```{r}
newdata <- data

replaceNA <- function(x){ 
      averageActivityPerDay[as.character(x)]
      }

newdata[boolNaIndices,]$steps <- sapply(data[boolNaIndices,]$interval, replaceNA)
```
### What is mean total number of steps taken per day in the new data?
```{r}
### Computing the median and mean of total number of steps per day in the new 
### data
## We use the previously defined totalSteps function and then compute the mean
## and median and construt a histogram to show the frequency count of total step
## in the filled in data.
totalStepValues <- totalSteps(newdata)
hist(totalStepValues)
mean(totalStepValues)
median(totalStepValues)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
weekDay_or_weekEnd <- function(x){
      dayOfX <- weekdays(x)
      typeOfDay <- character(length(x))
      typeOfDay[dayOfX=="Sunday"|dayOfX=="Saturday"] = "weekend"
      typeOfDay[!(dayOfX=="Sunday"|dayOfX=="Saturday")] = "weekday"
      return(typeOfDay)
}
newdata$weekday_or_weekend <- weekDay_or_weekEnd(data$date)

## Separates the data by weekday vs. weekend
splitData <- split(newdata, newdata$weekday_or_weekend)

## Computes the activity pattern separately for weekdays and weekends
weekendActivity <- activityPattern(splitData[["weekend"]])
splitData[["weekend"]]$averageActivity <- weekendActivity$averageActivityPerDay

weekdayActivity <- activityPattern(splitData[["weekday"]])
splitData[["weekday"]]$averageActivity <- weekdayActivity$averageActivityPerDay

## Construct a panel plot showing the average activity time series for weekends
## vs. weekdays
par(mfrow=c(2,1))
with(splitData[["weekday"]],plot(timeOfDay,averageActivityPerDay, type ="l",xlab
                                 ="Time of Day",ylab="Average Steps per Day",
                                 main= "Weekday") )
with(splitData[["weekend"]],plot(timeOfDay,averageActivityPerDay, type ="l",xlab
                                 ="Time of Day",ylab="Average Steps per Day",
                                 main= "Weekend") )
```


## Comments about the results obtained from this analysis (Unrelated to assignment but fun).
The way that it is plotted hides small differences. So perhaps it is worth considering a plot that overlays the two time series.
```{r}
with(splitData[["weekday"]],plot(timeOfDay,averageActivityPerDay, type ="l",xlab
                                 ="Time of Day",ylab="Average Steps per Day",
                                 main= "Daily activity") )
with(splitData[["weekend"]],points(timeOfDay,averageActivityPerDay,col="blue") )
```

Supprisingly, it seems that this person has the same activity pattern over weekdays as they have over weekends. This could either be a funny coincidence or more likely most of the missing values are from the weekends. Since I replaced the missing values with the interval averages and weekends are likely to yield NA. So lets examine this hypothesis :D!

##
```{r}
data$weekday_or_weekend <- weekDay_or_weekEnd(data$date)
splitOldData <- split(data, data$weekday_or_weekend)
sum(is.na(splitOldData[["weekday"]]$steps))/length(splitOldData[["weekday"]]$steps)
sum(is.na(splitOldData[["weekend"]]$steps))/length(splitOldData[["weekend"]]$steps)
```

Amazing! The proportion of missing values is the same in both weekdays and weekends so its unlikely that the filling in proceedure biased the results! :D wow!